% WMT 2026 system description — Task 3 (error-free segment detection)
% Deadline 7 Aug 2026 (SoftConf: https://softconf.com/emnlp2026/wmt2026/)
% NOT anonymized (system papers). Compile on Overleaf (no local LaTeX on the
% build machine): upload this directory as-is; pdfLaTeX + BibTeX.
% Numbers marked \todo{} must be regenerated from runs/ before submission.
\documentclass[11pt]{article}
\usepackage[final]{acl}
\usepackage{times}
\usepackage{latexsym}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{amsmath}
\newcommand{\todo}[1]{\textbf{\textcolor{red}{[TODO: #1]}}}
\usepackage{xcolor}

\title{Laptop-Scale Calibration Beats Bigger Models Alone:\\
Error-Free Segment Detection at the WMT26 Quality Evaluation Task}

\author{Ogulcan Sarioglu \\
  Independent Researcher \\
  \texttt{ogulcansarioglu1@gmail.com}}

\begin{document}
\maketitle

\begin{abstract}
We describe our submission to Task 3 (detection of error-free segments) of
the WMT26 Automated Translation Quality Evaluation shared task, with Task 2
as a by-product. Our system runs entirely on a single consumer laptop
(Apple M4 Max, MPS backend) at zero cloud cost: reference-free COMET-family
encoders (CometKiwi, xCOMET-XL) feed a transparent logistic-regression
ensemble over model scores, predicted-error-span mass, and segment length,
thresholded per language pair with MCC-optimal cutoffs and a guarded global
fallback for uncovered pairs. On a 54{,}419-segment development set built
from the WMT25 ESA human evaluation, per-language-pair calibration improves
macro MCC by ~70\% over a single global threshold (0.102 $\to$ 0.173 for
CometKiwi), the ensemble reaches pooled MCC \todo{0.29--0.33 final} versus
0.17 for the best single model, and — contrary to the received wisdom that
motivated this year's long-segment design — model-side truncation is
statistically indistinguishable from aligned chunked scoring for
regression-style QE at the observed 8.8\% over-limit exposure.
\todo{Official test results in camera-ready.}
\end{abstract}

\section{Introduction}
% Task 3: binary error-free detection, MCC-ranked; 23 LPs; long multi-sentence
% segments; references optional -> we are reference-free throughout.
\todo{1--2 paragraphs: task definition, cite task overview; our three design
commitments: (1) fully local/open (consumer laptop, \$0 cloud), (2)
calibration-centric rather than model-centric, (3) honest negative results.}

\section{System description}
\subsection{Base scorers}
CometKiwi \cite{rei-etal-2022-cometkiwi} and xCOMET-XL
\cite{guerreiro-etal-2024-xcomet}, reference-free inputs only, fp32 on MPS.
Throughput on one M4 Max: $\sim$10 seg/s (0.5B CometKiwi, batch 16),
$\sim$2 seg/s (3.5B xCOMET-XL, batch 4); the full 54k-segment dev sweep is
an overnight job.

\subsection{Long-segment handling}
Aligned mass-balanced chunking under the 512-token joint window; per-chunk
scores aggregated by min/mean/weighted-mean (min wins for Task 3 on dev:
pooled MCC 0.154 vs.\ 0.144). Unsplittable chunks are flagged, never hidden.
\todo{one paragraph algorithm sketch; refer to repo.}

\subsection{Ensemble}
Logistic regression over [kiwi score, XL score, severity/confidence-weighted
span mass, log length, LP one-hots]; class-balanced; unseen LPs fall back to
the base operating point (all-zero one-hots). Coefficients are published as
JSON — the entire decision function is auditable.

\subsection{Calibration protocol}
Per-LP MCC-optimal thresholds on an 80/20 stratified split; global fallback
when the fit split is small, one-class, or the per-LP threshold degenerates
to single-class predictions on held-out data. Prediction rule: score $>$
threshold $\Rightarrow$ error-free.

\section{Development setup}
Dev set: WMT25 General-MT ESA human evaluation \cite{kocmi-etal-2025-findings},
converted to 54{,}419 (segment, system) pairs over 14 LPs; gold label =
mean ESA $\geq 85$ AND zero annotated spans, mirroring the organizers'
documented derivation. Global error-free rate 0.134; per-LP 0.00--0.39.
10 of 23 WMT26 pairs have direct dev coverage; the rest ride the global
threshold — a transfer setting we analyze explicitly.

\section{Results on development data}
\subsection{Per-LP vs.\ global thresholds}
\todo{table from runs/dev/calibration/*/results.csv; headline: macro MCC
0.102 (global) vs 0.173 (per-LP) for CometKiwi; per-LP wins 9/13; global
threshold predicts 47--74\% positive on pairs whose true rate is 3--25\%.}

\subsection{Truncation vs.\ chunking (negative result)}
\todo{table: corpus-level wash (0.173 vs 0.169 macro), STILL a wash on the
4{,}791 over-limit segments (0.169 vs 0.176); interpretation: a 512-token
prefix is a sufficient quality sample for a regression scorer; caveat: test
exposure re-measured on 23 July; bootstrap CIs.}

\subsection{Low-resource behavior}
\todo{score distributions (fig), en-ar\_EG inversion case study, threshold
transfer quality for uncovered pairs, en-mas\_KE degenerate pair.}

\subsection{Model comparison and ensemble}
\todo{table: kiwi vs XL vs ensemble on identical splits, 4-seed holdout
(pooled 0.29--0.33; macro 0.21--0.23); xCOMET zero-span rule MCC $\approx$ 0
but span mass is a useful ensemble feature; feature ablation table; Task 2
Spearman by model.}

\section{Official results}
\todo{camera-ready: leaderboard placement + post-hoc analysis.}

\section{Conclusion}
\todo{3 sentences: calibration and honest evaluation beat raw scale at fixed
(laptop) compute; everything reproducible from the public repo.}

\section*{Limitations}
\todo{sentence-splitter simplicity; single-machine wall-clock; dev-test
distribution shift (threshold sensitivity of the gold rule); 2-variant
submission cap constrains what is officially scored.}

\bibliography{refs}

\end{document}
